|

### 2023-02-14 paper.tex
Tue Feb 14 13:00:08 CET 2023: update-remote.sh
Tue Feb 14 20:00:07 CET 2023: update-remote.sh

Johannes Gutenberg University Mainz}
\author{}
\date{\vspace{-2.5em}Date of Submission: 2023-02-13}

@@

Denoising Autoencoder Genetic Programming (DAE-GP) is a novel variation of an genetic programming based Estimation of Distribution Algorithm (EDA-GP) that uses a denoising autoencoders long short-term memory network (DAE-LSTMs) as a probabilistic model to sample new candidate solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]}.

DAE-LSTMs are artificial neural networks that can be trained in an unsupervised learning environment to minimize a reconstruction error for encoding input data into a compressed representation and subsequently decoding the compressed representation back to the input dimension.
In DAE-GP, DAE-LSTMs are trained with a subset of high-fitness solutions selected from a parent population with the aim to capture their promising qualities.
The resulting model is then used to sample new offspring solutions by propagating partially mutated solutions from the parent population through the DAE-LSTMS {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]}.
In previous work DAE-GP has been shown to outperform GP for both a generalized version of the royal tree problem {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]} as well as for a real-world symbolic regression problem {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{35}{]}.

The DAE-GP algorithm first described by {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]} trains a DAE-LSTMs for each generation \(g\) of the search from scratch.
{[}\protect\hyperlink{ref-dae-gp_2022_symreg}{35}{]} and {[}\protect\hyperlink{ref-daegp_explore_exploit}{34}{]} suggest the incorporation of a pre-training strategy into the evolutionary search as a possible way of improving the overall performance of the DAE-GP algorithm.
The key idea is to pre-train an initial DAE-LSTM on a large population of candidate solutions and to use the pre-trained parameters of this initial model in each generation of the search as starting parameters for the current generation DAE-LSTM.
This thesis studies the influence of using pre-trained DAE-LSTMs in DAE-GP for symbolic regression, especially looking at the influence on overall quality of solutions found by DAE-GP and effects on run-time.
The aim of this study is to answer the question if a pre-training strategy can be used to improve DAE-GP and to study both positive and negative effects on overall performance.

@@

DAE-GP is an EDA-GP algorithm that uses DAE-LSTM networks as a probabilistic model to sample new offspring solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]}.

@@

As a variant of GP, DAE-GP can be classified as a meta-heuristic that belongs to the field of evolutionary computation (EC).
EC based meta-heuristics are optimization methods that simulate the process of Darwinian evolution to search for high quality solutions by applying selection and variation operators to a population of candidate solutions.
Examples of EC include genetic algorithms (GA), evolutionary strategies (ES) and GP.
In EC, the quality of a solution is commonly measured as fitness and the time steps of the search are called generations.
Another important concept in EC is the distinction between genotypes and phenotypes of solutions.
The genotype contains the information that is necessary to construct the phenotype, the outer appearance of a particular solution on which we measure the overall quality of solutions.
The representation of a solution is therefore defined by the mapping of genotypes to phenotypes {[}\protect\hyperlink{ref-design_of_modern_heuristics}{23}{]}.
Genetic operators, such as mutation or recombination, are usually applied to the genotype of solutions.

@@

GP follows the same basic evolutionary principle of EC but searches for more general, hierarchical computer programs of dynamically varying size and shape {[}\protect\hyperlink{ref-Koza1993GeneticP}{15}{]}.
The computer programs that are at the center of the evolutionary search in GP are commonly represented by tree structures at the level of their phenotype {[}\protect\hyperlink{ref-design_of_modern_heuristics}{23}{]}.
Since GP searches for high fitness computer programs that produce a desired output for some input, it can be applied to various different problem domains such as symbolic regression, automatic programming, or evolving game-playing strategies {[}\protect\hyperlink{ref-Koza1993GeneticP}{15}{]}.
An important quality of GP is the ability to search for solutions of variable length and structure.
GP is an especially useful meta-heuristic for problems where no a priori knowledge about the final form of good solutions is available.
GPs ability to optimize solutions for both their structure as well as for their parameters led to it being one of the most prevalent methods used for symbolic regression {[}\protect\hyperlink{ref-10.1007ux2f978-3-540-24621-3_22}{20}{]}.

@@

The aim of estimation of distribution algorithms (EDA) is to replace the standard variation operators used in EC by building probabilistic models that can capture complex dependencies between different decision variables of an optimization problem {[}\protect\hyperlink{ref-design_of_modern_heuristics}{23}{]}.
EDAs use this probabilistic model to sample new offspring solutions inside an evolutionary search to replace crossover and/or mutation operators.
Even though a majority of EDA algorithms are designed to work with fixed length string representations as used in classical GAs, EDA algorithms have been researched and successfully applied to the domain of GP in various different research streams {[}\protect\hyperlink{ref-kim_probMod_GP}{13}{]}.

@@

One possible way of model building in EDA proposed by {[}\protect\hyperlink{ref-harmless_overfitting_eda}{22}{]} is to use denoising autoencoders (DAE) as generative models to capture complex probability distributions of decision variables.
DAE, a variation of the autoencoder (AE), is a widely used type of neural networks in the domain of unsupervised machine learning that maps \(n\) input variables to \(n\) output variables using a hidden representation.

@@

DAE was first introduced by {[}\protect\hyperlink{ref-dae_orig2008}{30}{]} as an improved AE with the ability to learn new representations of data that is especially robust to partially corrupted input data.
DAE modifies the AE by using partially corrupted input data for the AE and training it to reconstruct the uncorrupted, original version of the input data.

Since the hidden representation of DAE captures the dependency structure of the input data it can therefore also be used to generate new solutions in the context of GAs {[}\protect\hyperlink{ref-harmless_overfitting_eda}{22}{]}.

DAE-GP builds upon the concept of using DAEs in EDAs described by {[}\protect\hyperlink{ref-harmless_overfitting_eda}{22}{]} and transfers the concept to the domain of GP.
The mutation and crossover operators of standard GP are replaced by sampling new solutions from a probabilistic model that is build by training a DAEs long short-term memory (LSTM) network on a subset of high fitness solutions from the current population {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]}.

LSTMs are a variant of artificial neural networks first introduced by {[}\protect\hyperlink{ref-lstm_orig}{11}{]} that can store learned information over en extended period of time while avoiding the problem of vanishing and/or exploding gradients.
Since DAE-GP encodes candidate solutions as linear strings in prefix notation, the DAE in DAE-GP uses LSTMs for both encoding and decoding where the total amount of time steps \(T\) is equal to the sum of the length of the input solution and the output solution {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]}.

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM models \(M_g\) that are used in each DAE-GP generation \(g\) for \(g\in{1,...,g_{max}}\).
Instead of initializing and optimizing each model from scratch as done in previous work (e.g. {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]} {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{35}{]}), a separate DAE-LSTM network \(\hat{M}\) will be trained on a large population of randomly initialized solutions.\\

@@

Another motivation for introducing pre-training into DAE-GP is the prevalence of sampling error for small GP populations sizes.
{[}\protect\hyperlink{ref-sampling_err_gp}{24}{]} finds that sampling error, non-systematic errors that are caused by observing only a small subset of the statistical population, is a severe problem in the domain of GP since the initial population may not be a representative sample of all possible solutions.
{[}\protect\hyperlink{ref-sampling_err_gp}{24}{]} also introduces a method for calculating optimal population sizes to minimize the presence of sampling error.
Since DAE-LSTM of DAE-GP learns the properties of its training population and reuses the acquired knowledge in the sampling procedure, using the parameters obtained from training \(\hat{M}\) on a sufficiently large training population might reduce the need for large population sizes in the following generations of the search if \(\hat{M}\) already implicitly captured the properties of a representative sample of solutions.
If this mechanism can be applied successfully it would benefit the performance of DAE-GP by increasing the population diversity (resulting in better solution quality) as well as by reducing the need for large population sizes which require computational resources.

@@

The DAE-GP algorithm, first described by {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]}, is summarized as a flowchart in figure \ref{fig:daegp-flowchart}.
After setting the generation counter \(g\) to \(0\) and creating the initial population of solutions \(P_0\), DAE-GP enters the main loop that checks if a termination criteria is satisfied (i.e.~maximum number of generations or fitness evaluations is reached).
For each generation \(g\) the fitness of all new individuals in \(P_g\) is evaluated.
During model building, DAE-GP selects a subset \(X_g\) of the current population that is used as training data for the DAE-LSTM model \(M_g\) which is trained to learn the properties of \(X_g\).
The trained model \(M_g\) is then used for model sampling, partially mutated solutions from \(P_g\) are propagated through \(M_g\) to create a new population \(P_{g+1}\).
This process is repeated until the termination criteria is met and DAE-GP returns the highest fitness solutions.

@@

One difficulty in the implementation of a pre-training strategy into DAE-GP has been the determination of the number of hidden neurons inside the DAE-LSTMs hidden layers.
The original description of DAE-GP {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{36}{]} uses a strategy where the number of hidden neurons for each hidden layer is dynamically set per generation to the maximum individual size inside the current population.
For a pre-training implementation, this strategy can not be easily adapted since it leads to a changing number of neurons at each generation resulting in different dimensions of the DAE-LSTM.
To allow the sharing of the state of the pre-trained parameters \(\theta_{\hat{M}}\) from the pre-trained model \(\hat{M}\) to each \(M_g\) for \(g\in{1,...,g_{max}}\) the number of hidden neurons per hidden layer for all DAE-GP runs, pre-trained as well as regular, is statically set to a predefined value.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5116}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4884}}@{}}
\caption{\label{tab:baseParams-DAEGP}DAE-LSTM - Hyperparameter}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Hyperparameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Hyperparameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Stopping Condition & Convergence {[}\protect\hyperlink{ref-daegp_explore_exploit}{34}{]} \\
Reconstruction Error Metric & Multiclass Cross Entropy \\
Gradient Descent Algorithm & Adam Optimization {[}\protect\hyperlink{ref-Kingma2014AdamAM}{14}{]} \\
Learning Rate & 0.001 \\
Batch Size & 0.1 \\
Sampling Steps & 1 \\
Hidden Layer & 2 \\
Hidden Neurons & 150 \\
Corruption Operator & Levenshtein Edit {[}\protect\hyperlink{ref-daegp_explore_exploit}{34}{]} \\
Denoising Edit Probability (Model Building) & 0.05 \\
Denoising Edit Probability (Model Sampling) & 0.95 \\
Pre-Training Stopping Condition & Early Stopping \\
\end{longtable}

The framework that was used to run all experiments of this thesis was provided by the supervisor of this thesis David Wittenberg and uses the python programming language {[}\protect\hyperlink{ref-python3}{28}{]} in conjunction with the baseline libraries \texttt{Keras} {[}\protect\hyperlink{ref-chollet2015keras}{2}{]} for deep neural networks and \texttt{deap} {[}\protect\hyperlink{ref-DEAP_JMLR2012}{6}{]} for evolutionary computation.
For data visualizations I used the python package \texttt{matplotlib} {[}\protect\hyperlink{ref-matplotlib}{12}{]}, statistical analysis and evaluations were conducted using primarily the data science libraries \texttt{scipy} {[}\protect\hyperlink{ref-scipy}{31}{]}, \texttt{numpy} {[}\protect\hyperlink{ref-numpy}{8}{]}, \texttt{pandas} {[}\protect\hyperlink{ref-pandas}{26}{]}, and \texttt{IPython} {[}\protect\hyperlink{ref-ipython}{21}{]}.

@@

The objective of the airfoil problem is to find a function that accurately predicts the output variable \(y\) by taking in a subset of the input variables \(x1,x2,x3,x4,x5\).
The function set used by all DAE-GP variations for symbolic regression is summarized in tables \ref{tab:baseParams-GP}.
The terminal set consists of the the input variables \(x1,x2,x3,x4,x5\) and the ephemeral integer constants listed in table \ref{tab:baseParams-GP} {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{35}{]}.

@@

To gather a deeper understanding about the effect of pre-training DAE-LSTM in DAE-GP a series of experiments was conducted using the airfoil dataset for symbolic regression while using different parameter configurations for the number of hidden layers as well as the number of hidden neurons per hidden layer.

To study generalization behavior this series of experiments is conducted using DAE-GP with only a single generation until the search terminates.
I disregard the fitness of solutions found and focus solely on the reconstruction error that is produced during the training of each DAE-LSTM.
The reconstruction error is measured for two separate populations, a training population \(P_{train}\) that is used to train our DAE-LSTM as well as a hold-out validation population \(P_{test}\).
For the pre-trained DAE-GP two additional, separate populations \(\hat{P}_{train}\) and \(\hat{P}_{test}\) are used exclusively for pre-training.

@@

For each configuration I tested traditional DAE-GP as well as a pre-trained DAE-GP resulting in 12 total sub experiments that were each based on 10 individual runs (total number of runs=\(120\)).
To avoid creating biased results through the presence of sampling error inside the pre-training population (see {[}\protect\hyperlink{ref-sampling_err_gp}{24}{]}), the population size for the pre-training phase is chosen very high with the size of \(\hat{P}\) =10000 where 50\% of \(\hat{P}\) is used for the training population \(\hat{P}_{train}\) and the remaining 50\% are used for the hold-out validation population \(\hat{P}_{test}\).
The training of DAE-LSTM uses a fixed number of 1000 epochs until termination and uses Adam optimization for gradient descent.
The reason for using a high amount of 1000 fixed training epochs is to deliberately force the DAE-LSTM to overfit to the training data.

In general I expect that with a growing number of trainable parameters (either by adding more hidden layers or more hidden neurons per layer) the DAE-LSTM will be more prone to overfit to the training population \(P_{train}\) resulting in a small reconstruction error for the training population and a large one for the validation population \(P_{test}\).
The reason for this effect is that a larger network trained over an extended period of time (without the use of strategies like early stopping), has much more potential to learn noise from the training dataset than a smaller network, which is more likely to result in worsening performance for previously unseen cases {[}\protect\hyperlink{ref-weigend1994overfitting}{32}{]}.

Based on the review of {[}\protect\hyperlink{ref-pmlr-v5-erhan09a}{5}{]} I also expect that pre-training will have an insignificant or even negative influence on small DAE-LSTM instances while improving the networks generalization ability with growing size.

@@

The results for the experiment with varying number of hidden layers are summarized in figure \ref{fig:first-gen-airfoil-byLayers}.
Again, it can be observed that by increasing the dimension of the DAE-LSTM network both algorithms increasingly overfit to the training data.
While for a single hidden layer overfitting begins at around the 400th.
epoch, doubling the number of hidden layers to two leads to both algorithms overfitting at around the 200.th epoch.
In general over all three sub experiments, both algorithms start to overfit after approximately the same number of epochs but pre-training leads to reaching the point of overfitting at a higher reconstruction error which would result in a inferior model performance if the early stopping training strategy would be used during model training.

@@

Figure \ref{fig:airfoil-median-fitness-30gens} shows both the mean and median of the best fitness for each generation on both the test and training set of the airfoil dataset for using 2 hidden layers inside the DAE-LSTM networks.
The results are aggregated for \(10\) individuals runs per algorithm and also include regular GP as a benchmark.
For all three algorithms an improvement in the best found fitness can be observed during the evolutionary search but the results show that DAE-GP based algorithms, on average, find solutions with much higher fitness as traditional GP.

The first interesting observation from this experiment is, that the pre-trained DAE-GP shows slightly lower mean RMSE (higher fitness) but a higher median RMSE (lower fitness) than regular DAE-GP for both the training and the testing set in overall fitness.
Regarding the presence of overfitting, both DAE-GP variants show only a very small gap between the testing and training fitness which indicates a good generalization behavior.
Another interesting observation from figure \ref{fig:airfoil-median-fitness-30gens} is that pre-trained DAE-GP is faster at improving fitness than regular DAE-GP, this is especially true if only looking at the first 10 generations.
The median best RMSE on the test set after 10 generations for pre-trained DAE-GP is 17.093 which is smaller by a factor of 0.558 than regular DAE-GP with a median best RMSE of 30.658

@@

As expected, this reduction of hidden neurons by dynamic adjustment, does also result in a reduction of the fitness of solutions found.
Figure \ref{fig:airfoil-maxIndSize-fitness} shows that pre-trained DAE-GP, with its more complex DAE-LSTMs over the full 30 generations, does achieve a higher median fitness than regular DAE-GP with dynamic adjustment of hidden neurons.
Pre-Training does therefore introduce a new Trade-Off decision into the adjustment of DAE-GP hyperparameters where the number of hidden neurons needs to be carefully selected since it positively impacts solution quality but also increases the computational cost of DAE-LSTM training.

>>>>

Johannes Gutenberg University Mainz}
\author{}
\date{\vspace{-2.5em}Date of Submission: 2023-02-14}

@@

Denoising Autoencoder Genetic Programming (DAE-GP) is a novel variation of an genetic programming based Estimation of Distribution Algorithm (EDA-GP) that uses a denoising autoencoders long short-term memory network (DAE-LSTMs) as a probabilistic model to sample new candidate solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]}.

DAE-LSTMs are artificial neural networks that can be trained in an unsupervised learning environment to minimize a reconstruction error for encoding input data into a compressed representation and subsequently decoding the compressed representation back to the input dimension.
In DAE-GP, DAE-LSTMs are trained with a subset of high-fitness solutions selected from a parent population with the aim to capture their promising qualities.
The resulting model is then used to sample new offspring solutions by propagating partially mutated solutions from the parent population through the DAE-LSTMS {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]}.
In previous work DAE-GP has been shown to outperform GP for both a generalized version of the royal tree problem {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]} as well as for a real-world symbolic regression problem {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{34}{]}.

The DAE-GP algorithm first described by {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]} trains a DAE-LSTMs for each generation \(g\) of the search from scratch.
{[}\protect\hyperlink{ref-dae-gp_2022_symreg}{34}{]} and {[}\protect\hyperlink{ref-daegp_explore_exploit}{33}{]} suggest the incorporation of a pre-training strategy into the evolutionary search as a possible way of improving the overall performance of the DAE-GP algorithm.
The key idea is to pre-train an initial DAE-LSTM on a large population of candidate solutions and to use the pre-trained parameters of this initial model in each generation of the search as starting parameters for the current generation DAE-LSTM.
This thesis studies the influence of using pre-trained DAE-LSTMs in DAE-GP for symbolic regression, especially looking at the influence on overall quality of solutions found by DAE-GP and effects on run-time.
The aim of this study is to answer the question if a pre-training strategy can be used to improve DAE-GP and to study both positive and negative effects on overall performance.

@@

DAE-GP is an EDA-GP algorithm that uses DAE-LSTM networks as a probabilistic model to sample new offspring solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]}.

@@

As a variant of GP, DAE-GP can be classified as a meta-heuristic that belongs to the field of evolutionary computation (EC).
EC based meta-heuristics are optimization methods that simulate the process of Darwinian evolution to search for high quality solutions by applying selection and variation operators to a population of candidate solutions.
Examples of EC include genetic algorithms (GA), evolutionary strategies (ES) and GP.
In EC, the quality of a solution is commonly measured as fitness and the time steps of the search are called generations.
Another important concept in EC is the distinction between genotypes and phenotypes of solutions.
The genotype contains the information that is necessary to construct the phenotype, the outer appearance of a particular solution on which we measure the overall quality of solutions.
The representation of a solution is therefore defined by the mapping of genotypes to phenotypes {[}\protect\hyperlink{ref-design_of_modern_heuristics}{22}{]}.
Genetic operators, such as mutation or recombination, are usually applied to the genotype of solutions.

@@

GP follows the same basic evolutionary principle of EC but searches for more general, hierarchical computer programs of dynamically varying size and shape {[}\protect\hyperlink{ref-Koza1993GeneticP}{15}{]}.
The computer programs that are at the center of the evolutionary search in GP are commonly represented by tree structures at the level of their phenotype {[}\protect\hyperlink{ref-design_of_modern_heuristics}{22}{]}.
Since GP searches for high fitness computer programs that produce a desired output for some input, it can be applied to various different problem domains such as symbolic regression, automatic programming, or evolving game-playing strategies {[}\protect\hyperlink{ref-Koza1993GeneticP}{15}{]}.
An important quality of GP is the ability to search for solutions of variable length and structure.
GP is an especially useful meta-heuristic for problems where no a priori knowledge about the final form of good solutions is available.
GPs ability to optimize solutions for both their structure as well as for their parameters led to it being one of the most prevalent methods used for symbolic regression {[}\protect\hyperlink{ref-10.1007ux2f978-3-540-24621-3_22}{20}{]}.

@@

The aim of estimation of distribution algorithms (EDA) is to replace the standard variation operators used in EC by building probabilistic models that can capture complex dependencies between different decision variables of an optimization problem {[}\protect\hyperlink{ref-design_of_modern_heuristics}{22}{]}.
EDAs use this probabilistic model to sample new offspring solutions inside an evolutionary search to replace crossover and/or mutation operators.
Even though a majority of EDA algorithms are designed to work with fixed length string representations as used in classical GAs, EDA algorithms have been researched and successfully applied to the domain of GP in various different research streams {[}\protect\hyperlink{ref-kim_probMod_GP}{13}{]}.

@@

One possible way of model building in EDA proposed by {[}\protect\hyperlink{ref-harmless_overfitting_eda}{21}{]} is to use denoising autoencoders (DAE) as generative models to capture complex probability distributions of decision variables.
DAE, a variation of the autoencoder (AE), is a widely used type of neural networks in the domain of unsupervised machine learning that maps \(n\) input variables to \(n\) output variables using a hidden representation.

@@

DAE was first introduced by {[}\protect\hyperlink{ref-dae_orig2008}{29}{]} as an improved AE with the ability to learn new representations of data that is especially robust to partially corrupted input data.
DAE modifies the AE by using partially corrupted input data for the AE and training it to reconstruct the uncorrupted, original version of the input data.

Since the hidden representation of DAE captures the dependency structure of the input data it can therefore also be used to generate new solutions in the context of GAs {[}\protect\hyperlink{ref-harmless_overfitting_eda}{21}{]}.

DAE-GP builds upon the concept of using DAEs in EDAs described by {[}\protect\hyperlink{ref-harmless_overfitting_eda}{21}{]} and transfers the concept to the domain of GP.
The mutation and crossover operators of standard GP are replaced by sampling new solutions from a probabilistic model that is build by training a DAEs long short-term memory (LSTM) network on a subset of high fitness solutions from the current population {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]}.

LSTMs are a variant of artificial neural networks first introduced by {[}\protect\hyperlink{ref-lstm_orig}{11}{]} that can store learned information over en extended period of time while avoiding the problem of vanishing and/or exploding gradients.
Since DAE-GP encodes candidate solutions as linear strings in prefix notation, the DAE in DAE-GP uses LSTMs for both encoding and decoding where the total amount of time steps \(T\) is equal to the sum of the length of the input solution and the output solution {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]}.

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM models \(M_g\) that are used in each DAE-GP generation \(g\) for \(g\in{1,...,g_{max}}\).
Instead of initializing and optimizing each model from scratch as done in previous work (e.g. {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]} {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{34}{]}), a separate DAE-LSTM network \(\hat{M}\) will be trained on a large population of randomly initialized solutions.\\

@@

Another motivation for introducing pre-training into DAE-GP is the prevalence of sampling error for small GP populations sizes.
{[}\protect\hyperlink{ref-sampling_err_gp}{23}{]} finds that sampling error, non-systematic errors that are caused by observing only a small subset of the statistical population, is a severe problem in the domain of GP since the initial population may not be a representative sample of all possible solutions.
{[}\protect\hyperlink{ref-sampling_err_gp}{23}{]} also introduces a method for calculating optimal population sizes to minimize the presence of sampling error.
Since DAE-LSTM of DAE-GP learns the properties of its training population and reuses the acquired knowledge in the sampling procedure, using the parameters obtained from training \(\hat{M}\) on a sufficiently large training population might reduce the need for large population sizes in the following generations of the search if \(\hat{M}\) already implicitly captured the properties of a representative sample of solutions.
If this mechanism can be applied successfully it would benefit the performance of DAE-GP by increasing the population diversity (resulting in better solution quality) as well as by reducing the need for large population sizes which require computational resources.

@@

The DAE-GP algorithm, first described by {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]}, is summarized as a flowchart in figure \ref{fig:daegp-flowchart}.
After setting the generation counter \(g\) to \(0\) and creating the initial population of solutions \(P_0\), DAE-GP enters the main loop that checks if a termination criteria is satisfied (i.e.~maximum number of generations or fitness evaluations is reached).
For each generation \(g\) the fitness of all new individuals in \(P_g\) is evaluated.
During model building, DAE-GP selects a subset \(X_g\) of the current population that is used as training data for the DAE-LSTM model \(M_g\) which is trained to learn the properties of \(X_g\).
The trained model \(M_g\) is then used for model sampling, partially mutated solutions from \(P_g\) are propagated through \(M_g\) to create a new population \(P_{g+1}\).
This process is repeated until the termination criteria is met and DAE-GP returns the highest fitness solutions.

@@

One difficulty in the implementation of a pre-training strategy into DAE-GP has been the determination of the number of hidden neurons inside the DAE-LSTMs hidden layers.
The original description of DAE-GP {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{35}{]} uses a strategy where the number of hidden neurons for each hidden layer is dynamically set per generation to the maximum individual size inside the current population.
For a pre-training implementation, this strategy can not be easily adapted since it leads to a changing number of neurons at each generation resulting in different dimensions of the DAE-LSTM.
To allow the sharing of the state of the pre-trained parameters \(\theta_{\hat{M}}\) from the pre-trained model \(\hat{M}\) to each \(M_g\) for \(g\in{1,...,g_{max}}\) the number of hidden neurons per hidden layer for all DAE-GP runs, pre-trained as well as regular, is statically set to a predefined value.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5116}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4884}}@{}}
\caption{\label{tab:baseParams-DAEGP}DAE-LSTM - Hyperparameter}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Hyperparameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Hyperparameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Stopping Condition & Convergence {[}\protect\hyperlink{ref-daegp_explore_exploit}{33}{]} \\
Reconstruction Error Metric & Multiclass Cross Entropy \\
Gradient Descent Algorithm & Adam Optimization {[}\protect\hyperlink{ref-Kingma2014AdamAM}{14}{]} \\
Learning Rate & 0.001 \\
Batch Size & 0.1 \\
Sampling Steps & 1 \\
Hidden Layer & 2 \\
Hidden Neurons & 150 \\
Corruption Operator & Levenshtein Edit {[}\protect\hyperlink{ref-daegp_explore_exploit}{33}{]} \\
Denoising Edit Probability (Model Building) & 0.05 \\
Denoising Edit Probability (Model Sampling) & 0.95 \\
Pre-Training Stopping Condition & Early Stopping \\
\end{longtable}

The framework that was used to run all experiments of this thesis was provided by the supervisor of this thesis David Wittenberg and uses the python programming language {[}\protect\hyperlink{ref-python3}{27}{]} in conjunction with the baseline libraries \texttt{Keras} {[}\protect\hyperlink{ref-chollet2015keras}{2}{]} for deep neural networks and \texttt{deap} {[}\protect\hyperlink{ref-DEAP_JMLR2012}{6}{]} for evolutionary computation.
For data visualizations the python package \texttt{matplotlib} {[}\protect\hyperlink{ref-matplotlib}{12}{]} was used, statistical analysis and evaluations were conducted using primarily the data science libraries \texttt{scipy} {[}\protect\hyperlink{ref-scipy}{30}{]}, \texttt{numpy} {[}\protect\hyperlink{ref-numpy}{8}{]} and \texttt{pandas} {[}\protect\hyperlink{ref-pandas}{25}{]}.

@@

The objective of the airfoil problem is to find a function that accurately predicts the output variable \(y\) by taking in a subset of the input variables \(x1,x2,x3,x4,x5\).
The function set used by all DAE-GP variations for symbolic regression is summarized in tables \ref{tab:baseParams-GP}.
The terminal set consists of the the input variables \(x1,x2,x3,x4,x5\) and the ephemeral integer constants listed in table \ref{tab:baseParams-GP} {[}\protect\hyperlink{ref-dae-gp_2022_symreg}{34}{]}.

@@

To gain a deeper understanding about the effect of pre-training the DAE-LSTM networks in DAE-GP, a series of experiments was conducted using the airfoil dataset for symbolic regression while using different parameter configurations for the number of hidden layers as well as the number of hidden neurons per hidden layer.

To study the generalization behavior of the DAE-LSTM model, this series of experiments is conducted using DAE-GP with only a single generation until the search terminates.
The fitness of solutions found was disregarded to focus solely on the reconstruction error that is produced during the training of each DAE-LSTM.
The reconstruction error is measured for two separate populations, a training population \(P_{train}\) that is used to train our DAE-LSTM as well as a hold-out validation population \(P_{test}\).
For the pre-trained DAE-GP two additional, separate populations \(\hat{P}_{train}\) and \(\hat{P}_{test}\) are used exclusively for pre-training.

@@

For each configuration, traditional DAE-GP as well as a pre-trained DAE-GP was tested resulting in 12 total sub experiments that were each based on 10 individual runs (total number of runs=\(120\)).
To avoid creating biased results through the presence of sampling error inside the pre-training population (see {[}\protect\hyperlink{ref-sampling_err_gp}{23}{]}), the population size for the pre-training phase is chosen very high with the size of \(\hat{P}\) =10000 where 50\% of \(\hat{P}\) is used for the training population \(\hat{P}_{train}\) and the remaining 50\% are used for the hold-out validation population \(\hat{P}_{test}\).
The training of each DAE-LSTM stopped after a fixed number of 1000 epochs.
The reason for using a high amount of 1000 fixed training epochs was to deliberately force the DAE-LSTM to overfit to the training data.

In general, it is expected that with a growing number of trainable parameters (either by adding more hidden layers or more hidden neurons per layer) the DAE-LSTM will be more likely to overfit to the training population \(P_{train}\) which will result in a small reconstruction error for the training population and a large one for the validation population \(P_{test}\).
The reason for this effect is that a larger network trained over an extended period of time (without the use of strategies like early stopping), has much more potential to learn noise from the training dataset than a smaller network, which is more likely to result in worsening performance for previously unseen cases {[}\protect\hyperlink{ref-weigend1994overfitting}{31}{]}.

Based on the review of {[}\protect\hyperlink{ref-pmlr-v5-erhan09a}{5}{]} it is also expected that pre-training will have an insignificant or even negative influence on small DAE-LSTM instances while improving the networks generalization ability with growing size.

@@

The results for the experiment with a varying number of hidden layers are summarized in figure \ref{fig:first-gen-airfoil-byLayers}.
Again, it can be observed that by increasing the dimension of the DAE-LSTM network both algorithms increasingly overfit to the training data.
While for a single hidden layer overfitting begins at around the 400th.
epoch, doubling the number of hidden layers to two leads to both algorithms overfitting at around the 200.th epoch.
In general over all three sub experiments, both algorithms start to overfit after approximately the same number of epochs but pre-training leads to reaching the point of overfitting at a higher reconstruction error which would result in a inferior model performance if the early stopping training strategy would be used during model training.

@@

Figure \ref{fig:airfoil-median-fitness-30gens} shows both the mean and median the best fitness by generations for the airfoil dataset.
The results are aggregated for \(10\) individuals runs per algorithm and also include regular GP as a benchmark.
For all three algorithms an improvement in the best found fitness can be observed during the evolutionary search but the results show that DAE-GP based algorithms, on average, find solutions with much higher fitness as traditional GP.

The first interesting observation from this experiment is, that the pre-trained DAE-GP shows slightly lower mean RMSE (higher fitness) but a higher median RMSE (lower fitness) than regular DAE-GP for both the training and the testing set in overall fitness.
Regarding the presence of overfitting, both DAE-GP variants show only a very small gap between the testing and training fitness which indicates a good generalization behavior.\\

Another interesting observation from figure \ref{fig:airfoil-median-fitness-30gens} is that pre-trained DAE-GP is faster at improving fitness than regular DAE-GP.
The median best RMSE on the test set after 10 generations for pre-trained DAE-GP is 17.093 which is smaller by a factor of 0.558 than regular DAE-GP with a median best RMSE of 30.658

@@

As expected, the reduction of hidden neurons by dynamic adjustment does also result in a reduction of the fitness of solutions found.
Figure \ref{fig:airfoil-maxIndSize-fitness} shows that pre-trained DAE-GP, with its more complex DAE-LSTMs over the full 30 generations, does achieve a higher median fitness than regular DAE-GP with dynamic adjustment of hidden neurons.
Pre-Training does therefore introduce a new Trade-Off decision into the adjustment of DAE-GP hyperparameters where the number of hidden neurons needs to be carefully selected since it positively impacts solution quality but also increases the computational cost of DAE-LSTM training.
